Species turnover reveals hidden effects of decreasing Nitrogen deposition in mountain hay meadows

### 2018-06-28 Manuscript.Rmd
Update Manuscript.Rmd

The five measures of plant community structure (i.e. biodiversity endpoints) suggested that plant communities in mountain hay meadows were rather stable between 2003 and 2017 and did not show a clear increase or decrease over time (Table \@ref(tab:communitytrendstab)): for each of the three 5-year survey periods the averages of species-richness and the average indicator values for temperature, humidity, nutrients and light did not vary much among the three sampling periods and the estimated trends were rather small. Except for average indicator value for temperature, the 90% credible-interval of the temporal trend contained zero. The results from the linear mixed models suggest that a linear temporal change was most likely for the community mean of the indicator value for temperature (probability of increase: `r 
res %>% filter(Measures == "Temperature value") %>% pull("Prob. for Trend")`), followed by the community mean of the indicator value for light (probability of decrease: `r 1 - res %>% filter(Measures == "Light value") %>% pull("Prob. for Trend")`) and it was least likely for the species richness (probability of increase: `r res %>% filter(Measures == "Species-richness") %>% pull("Prob. for Trend")`). The chance that the community mean of the nutrient value decreased between 2003 and 2017 was `r 1- res %>% filter(Measures == "Nutrients value") %>% pull("Prob. for Trend")`.

@@

High species turnover at a site is the result of species that disappeared from the site and species that newly colonized the site. To better understand the factors that drive these changes we were particularly interested whether the species that disappeared or colonized the sites differed in indicator values compared to what would be expected if the same number of species randomly disappeared or colonized the sites (i.e. random disappearance and random colonization) and whether there is a spatial variation along the gradients. It seems that the indicator values of newly colonizing species differed more from random colonization than the indicator values of disappearing species (Table \@ref(tab:difffromrandomtab)). For colonizing species, we found the largest differences from random colonization in the indicator value for nutrients: at sites with nitrogen deposition of 10 kg ha$^{-1}$ yr$^{-1}$ the newly colonizing species had in average a lower indicator value for nutrients than species under random disappearance (column "Difference from random" in Table \@ref(tab:difffromrandomtab)), but this differences between colonizing species and random colonization decreased with increasing Nitrogen deposition (column "Change along gradient" in Table \@ref(tab:difffromrandomtab)). Thus, at high Nitrogen deposition colonizing species did not differ from random species (see Figure 3b in Appendix A).

While colonizing species had higher temperature values compared to what we would expect under random colonization, the differences between colonizing species and random colonization was about four times smaller compared to the difference in indicator value for nutrients between colonizing species and random species. Nevertheless, the variation in indicator value for temperature seemed important to explain the total species turnover. This is because, disappearing species tend to have lower temperature value than random species as well as colonizing species tend to have higher temperature values than random species; both processes lead to an overall replacement of species with lower temperature value with species with higher temperature values. This was not the case for the indicator value for nutrients: species with lower nutrients values tended to be more likely to disappear from as well as to colonize sites compared to random species (Table \@ref(tab:difffromrandomtab)). See also Appendix A where we present detailed results for the comparison between colonizing or disappearing species with randomly selected species.

@@

Nitrogen deposition decreased with increasing elevation (Fig. \@ref(fig:Ndepchange)a). In 2000 only `r 100 * mean(sites$NTOT_2000 < 10)`% of sites had a Nitrogen deposition rate of less than 10 kg N ha$^{-1}$ yr$^{-1}$, which lied all above 1000m. Between 2000 and 2015 the Nitrogen deposition decreased in average `r mean(sites$NTOT_2015 - sites$NTOT_2000)` $\pm$ `r sd(sites$NTOT_2015 - sites$NTOT_2000)` kg N ha$^{-1}$ yr$^{-1}$ with slightly higher net decreases at lower elevation (Fig. \@ref(fig:Ndepchange)a).

@@

While we could not detect a consistent decrease in average indicator value for nutrients (Table \@ref(tab:communitytrendstab)), the higher colonization rate of species with low nutrient value at sites with low deposition rate seems to affect the spatial variation of oligotrophic species richness: sites with low Nitrogen deposition are likely to become more species rich over time likely resulting in steeper slope of the negative relationship between Nitrogen deposition and oligotrophic species richness. Indeed, if we apply at different time points a similar model as in @Roth2013 to infer the effects of Nitrogen deposition on the spatial variation of oligotrophic species richness, the resulting effect size (i.e. the slope) became more negative over time (Fig. \@ref(fig:figconsequences)).

@@

# Discussion {-}
Although Nitrogen deposition considerably declined between 2000 and 2015 (Fig. \@ref(fig:Ndepchange)), we could detect only weak shifts in plant community structure (i.e. biodiversity endpoints sensu @Rowe2017) during the same time period (Table \@ref(tab:communitytrendstab)). While the slight increase in average temperature value suggests that plant communities adopted to increasing temperatures, the constant average nutrients value suggests that the decrease in Nitrogen deposition did not yet affect plant communities. However, this apparent stability in community structure was accompanied by marked temporal turnover in species identities. It seems unlikely that this temporal turnover can entirely be explained by methodological issues such as overlooked species. First, pseudo-turnover of species entities in independent surveys of the same site during the same season was smaller than the observed temporal turnover between two surveys from different years. Second, spatial variation of turnover showed patterns that can hardly be explained by methodological issues. For instance species turnover varied along the Nitrogen deposition gradient with highest species turnover at sites with low Nitrogen deposition (Table \@ref(tab:turnovertab)). Taken together, our results add to the increasing evidence that contemporary communities are often relatively stable at community level, but this apparent stability is often accompanied by marked turnover of species [@Vellend2013; @Dornelas2014; @Hillebrand2018].

Species communities are shaped by a range of factors, including deterministic processes such as environmental filtering or competitive interactions [@Goetzenberger2012; @Janevcek2013]. Such factors select for species with specific characteristics. Community assembly theories thus suggest that the factors that drive the composition of species in a community can be inferred from comparing the characteristics of the species in the community with random expectation [@Chase2011]. We adopted this idea and compared the indicator values of species that disappeared or colonized a site with the indicator values from randomly chosen species from the same site (Appendix A). We found that the nutrient values of colonization species showed largest deviations from random expectation suggesting that Nitrogen deposition or other factors that change the nutrient content of soils were the most important drivers of the change in species composition over the last 15 years in Swiss mountain hay meadows.

Nitrogen deposition at our plots decreased by `r mean(sites$NTOT_2015 - sites$NTOT_2000)` kg ha$^{-1}$ yr$^{-1}$ on average between 2000 and 2015. This is only about one tenth of the decrease in England, where Nitrogen deposition decreased by 24 kg ha$^{-1}$ yr$^{-1}$ from 1996 to 2011 [@Storkey2015]. This comparatively low change in Nitrogen deposition in Swiss mountain hay meadows might explain why we observed no change in average nutrient value of communities. Additionally, other anthropogenic pressures such as climate change might have overruled Nitrogen depositing effects on community structure. Particularly, we found that the species disappearing from the sites tend to have below average indicator values for temperature, while species that newly colonize sites have above average indicator values for temperature (Table \@ref(tab:difffromrandomtab)). Thus, the effect of disappearing and the effect of colonizing species on the community mean for temperature is additive resulting in increasing average temperature values (Tab. \@ref(tab:communitytrendstab)). This is in contrast to how Nitrogen deposition is affecting disappearing and colonization of species: It seems that both species disappearing from the sites as well as species colonizing the sites tend to have below average indicator values for nutrients (Table \@ref(tab:difffromrandomtab)). Thus, the effect of disappearing and colonizing  species on the average community value for nutrients partially cancel each other out. Furthermore, eutrophic species had rather high local survival across the entire deposition gradient, while oligotrophic species have much reduced local survival at higher Nitrogen deposition rates. This suggests that mountain hay meadow communities can reach alternative stable states with eutrophic species that are unlikely to disappear even if Nitrogen deposition is reduced [@Stevens2016]. Taken together, these reasons might explain why the response of mountain hay meadow communities responded stronger to climate warming than to Nitrogen reduction, although the reduction in Nitrogen resulted in above average colonization of oligotrophic species.

Observational studies along a gradient of Nitrogen deposition often infer how the spatial variation in species richness is related to Nitrogen deposition [@Stevens2010gradient; chapter 4 in @deVries2015]. They assume that the spatial variation in species richness (or other metrics of community structure) arose because over time some areas lost more species than others because they chronically experienced higher Nitrogen deposition. Although there is evidence supporting the use of such a 'space for time substitution' for detecting the effects of Nitrogen deposition on plant diversity [@Stevens2010], they can not replace studies that relate temporal patterns in species with Nitrogen deposition [@deSchrijver2011]. There are only a limited number of studies inferring temporal trends of plant species diversity related to varying amounts of Nitrogen deposition in extant communities [@Clark2008; @Storkey2015; @Stevens2016]. In a recent study, we used the mountain hay meadow data from a single survey only and estimated the critical load along the Nitrogen deposition gradient at which species richness of oligotrophic species richness start to decrease with increasing Nitrogen deposition [@Roth2017]. Using this spatial variation in species richness and nitrogen deposition we estimated a critical load for mountain hay meadows of 13.1 kg ha$^{-1}$ yr$^{-1}$. In the current study we estimated the rate of Nitrogen deposition at which local survival probability or colonization probability was equal for oligotrophic and eutrophic species. This rate could also be interpreted as a critical load. With this temporal approach we obtained very similar results as the spatial approach. Our results are thus validating the space for time approach, at least for Swiss mountain hay meadows.

>>>>

The five measures of plant community composition (i.e. biodiversity endpoints) suggested that plant communities in mountain hay meadows were rather stable between 2003 and 2017 and did not show a clear increase or decrease over time (Table \@ref(tab:communitytrendstab)): for each of the three 5-year survey periods the averages of species-richness and the average indicator values for temperature, humidity, nutrients and light did not vary much among the three sampling periods and the estimated trends were rather small. Except for average indicator value for temperature, the 90% credible-interval of the temporal trend contained zero. The results from the linear mixed models suggest that a linear temporal change was most likely for the community mean of the indicator value for temperature (probability of increase: `r 
res %>% filter(Measures == "Temperature value") %>% pull("Prob. for Trend")`), followed by the community mean of the indicator value for light (probability of decrease: `r 1 - res %>% filter(Measures == "Light value") %>% pull("Prob. for Trend")`). A linear temporal change was least likely for the species richness (probability of increase: `r res %>% filter(Measures == "Species-richness") %>% pull("Prob. for Trend")`). The chance that the community mean of the nutrient value decreased between 2003 and 2017 was `r 1- res %>% filter(Measures == "Nutrients value") %>% pull("Prob. for Trend")`.

@@

High species turnover at a site is the result of species that disappeared from the site and species that newly colonized the site. To better understand the factors that drive these changes we were particularly interested whether the species that disappeared or colonized the sites differed in indicator values compared to what would be expected if the same number of species randomly disappeared or colonized the sites (i.e. random disappearance and random colonization) and whether there is a spatial variation along the environmental gradients. It seems that the indicator values of newly colonizing species differed more from random colonization than the indicator values of disappearing species (Table \@ref(tab:difffromrandomtab)). For colonizing species, we found the largest differences from random colonization in the indicator value for nutrients: at sites with relatively low nitrogen deposition of 10 kg ha$^{-1}$ yr$^{-1}$ the newly colonizing species had in average a lower indicator value for nutrients than species under random disappearance (column "Difference from random" in Table \@ref(tab:difffromrandomtab)), but this differences between colonizing species and random colonization decreased with increasing Nitrogen deposition (column "Change along gradient" in Table \@ref(tab:difffromrandomtab)). Thus, at high Nitrogen deposition colonizing species did not differ from random species (see Figure 3b in Appendix A).

While colonizing species had higher temperature values compared to what we would expect under random colonization, the differences between colonizing species and random colonization was about four times smaller compared to the difference in indicator value for nutrients between colonizing species and random species. Nevertheless, the variation in indicator value for temperature seemed important to explain the total species turnover. This is because disappearing species tend to have lower temperature values than random species as well as colonizing species tend to have higher temperature values than random species; both processes lead to an overall replacement of species with lower temperature value with species with higher temperature values. This was not the case for the indicator value for nutrients: species with lower nutrients values tended to be more likely to disappear from as well as to colonize sites compared to random species (Table \@ref(tab:difffromrandomtab)). See also Appendix A where we present detailed results for the comparison between colonizing or disappearing species with randomly selected species.

@@

Nitrogen deposition decreased with increasing elevation (Fig. \@ref(fig:Ndepchange)a). In 2000 only `r 100 * mean(sites$NTOT_2000 < 10)`% of sites had a Nitrogen deposition rate of less than 10 kg N ha$^{-1}$ yr$^{-1}$. All of these sites are situated above 1000m. Between 2000 and 2015 the Nitrogen deposition decreased in average by `r mean(sites$NTOT_2015 - sites$NTOT_2000)` $\pm$ `r sd(sites$NTOT_2015 - sites$NTOT_2000)` kg N ha$^{-1}$ yr$^{-1}$ with slightly higher net decreases at lower elevation (Fig. \@ref(fig:Ndepchange)a).

@@

While we could not detect a consistent decrease in average indicator values for nutrients (Table \@ref(tab:communitytrendstab)), the higher colonization rate of species with low nutrient value at sites with low deposition rate seems to affect the spatial variation of oligotrophic species richness: sites with low Nitrogen deposition are likely to become more species rich over time likely resulting in steeper slope of the negative relationship between Nitrogen deposition and oligotrophic species richness. Indeed, if we apply a similar model as in @Roth2013 at different points in time to infer the effects of Nitrogen deposition on the spatial variation of oligotrophic species richness, the resulting effect size (i.e. the slope) became more negative over time (Fig. \@ref(fig:figconsequences)).

@@

# Discussion {-}
Although Nitrogen deposition considerably declined between 2000 and 2015 (Fig. \@ref(fig:Ndepchange)), we could detect only weak shifts in plant community structure (i.e. biodiversity endpoints sensu @Rowe2017) during the same time period (Table \@ref(tab:communitytrendstab)). While the slight increase in average temperature values  suggests that plant communities adopted to increasing temperatures, the constant average nutrients value suggests that the decrease in Nitrogen deposition did not yet affect plant communities. However, this apparent stability in community composition  was accompanied by marked temporal turnover in species identities. It seems unlikely that this temporal turnover can entirely be explained by methodological issues such as overlooked species. First, pseudo-turnover of species entities in independent surveys of the same site during the same season was smaller than the observed temporal turnover between two surveys from different years. Second, spatial variation of turnover showed patterns that can hardly be explained by methodological issues. For instance species turnover varied along the Nitrogen deposition gradient with highest species turnover at sites with low Nitrogen deposition (Table \@ref(tab:turnovertab)). Taken together, our results add to the increasing evidence that contemporary communities are often relatively stable at community level, but this apparent stability is often accompanied by marked turnover of species [@Vellend2013; @Dornelas2014; @Hillebrand2018].

Species communities are shaped by a range of factors, including deterministic processes such as environmental filtering or competitive interactions [@Goetzenberger2012; @Janevcek2013]. Such factors select for species with specific characteristics. Community assembly theories thus suggest that the factors that drive the composition of species in a community can be inferred from comparing the characteristics of the species in the community with random expectation [@Chase2011]. We adopted this idea and compared the indicator values of species that disappeared or colonized a site with the indicator values from randomly chosen species from the same site (Appendix A). We found that the nutrient values of colonizing species showed largest deviations from random expectation suggesting that Nitrogen deposition or other factors that change the nutrient content of soils were the most important drivers of the change in species composition over the last 15 years in Swiss mountain hay meadows.

Nitrogen deposition at our plots decreased by `r mean(sites$NTOT_2015 - sites$NTOT_2000)` kg ha$^{-1}$ yr$^{-1}$ on average between 2000 and 2015. This is only about one tenth of the decrease in England, where Nitrogen deposition decreased by 24 kg ha$^{-1}$ yr$^{-1}$ from 1996 to 2011 [@Storkey2015]. This comparatively low change in Nitrogen deposition in Swiss mountain hay meadows might explain why we observed no change in average nutrient value of communities. Additionally, other anthropogenic pressures such as climate change might have overruled Nitrogen depositing effects on community composition. Particularly, we found that the species disappearing from the sites tend to have below average indicator values for temperature, while species that newly colonize sites have above average indicator values for temperature (Table \@ref(tab:difffromrandomtab)). Thus, the effect of disappearing and the effect of colonizing species on the community mean for temperature is additive resulting in increasing average temperature values (Tab. \@ref(tab:communitytrendstab)). This is in contrast to how Nitrogen deposition is affecting disappearing and colonization of species: It seems that both species disappearing from the sites as well as species colonizing the sites tend to have below average indicator values for nutrients (Table \@ref(tab:difffromrandomtab)). Thus, the effect of disappearing and colonizing  species on the average community value for nutrients partially cancel each other out. Furthermore, eutrophic species had rather high local survival across the entire deposition gradient, while oligotrophic species have much reduced local survival at higher Nitrogen deposition rates. This suggests that mountain hay meadow communities can reach alternative stable states with eutrophic species that are unlikely to disappear even if Nitrogen deposition is reduced [@Stevens2016]. Taken together, these reasons might explain why the composition of mountain hay meadow communities responded stronger to climate warming than to Nitrogen reduction, although the reduction in Nitrogen resulted in above average colonization of oligotrophic species.

Observational studies along a gradient of Nitrogen deposition often infer how the spatial variation in species richness is related to Nitrogen deposition [@Stevens2010gradient; chapter 4 in @deVries2015]. They assume that the spatial variation in species richness (or other metrics of community composition) arose because over time some areas lost more species than others because they chronically experienced higher Nitrogen deposition. Although there is evidence supporting the use of such a 'space for time substitution' for detecting the effects of Nitrogen deposition on plant diversity [@Stevens2010], they can not replace studies that relate temporal patterns in species with Nitrogen deposition [@deSchrijver2011]. There are only a limited number of studies inferring temporal trends of plant species diversity related to varying amounts of Nitrogen deposition in extant communities [@Clark2008; @Storkey2015; @Stevens2016]. In a recent study, we used the mountain hay meadow data from a single survey only and estimated the critical load along the Nitrogen deposition gradient at which species richness of oligotrophic species richness start to decrease with increasing Nitrogen deposition [@Roth2017]. Using this spatial variation in species richness and nitrogen deposition we estimated a critical load for mountain hay meadows of 13.1 kg ha$^{-1}$ yr$^{-1}$. In the current study we estimated the rate of Nitrogen deposition at which local survival probability or colonization probability was equal for oligotrophic and eutrophic species. This rate could also be interpreted as a critical load. With this temporal approach we obtained very similar results as the spatial approach. Our results are thus validating the space for time approach, at least for Swiss mountain hay meadows.
